Making Higher-Order Superposition Work

### 2022-03-15 scratch/paper.tex
more

\section{Introduction}
\label{sec:ho-tech:intro}

@@

Many higher-order problems extensively use symbol definitions to simplify
their representation. We describe several ways to exploit the definitions,
%of which the most successful is
such as turning them into rewrite rules (Sect.~\ref{sec:ho-tech:preprocessing}).
%Interesting patterns can be observed in various higher-order problem encodings.
%We show how we can exploit these to simplify problems (Sect.~\ref{sec:ho-tech:preprocessing}).
%
By working on formulas rather than clauses, tableau techniques take a more
holistic view of a higher-order problem.
Through its support for delayed clausification and, more generally,
calculus-level formula manipulation, \osup{} enables us to
simulate most successful tableau techniques in a saturating prover
(Sect.~\ref{sec:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
functions}, a mechanism that allows us to choose on which Boolean subterms
to perform inferences first.
We implemented some Boolean selection functions and
evaluated them (Sect.~\ref{sec:ho-tech:bool-select}).

The main drawback of both \lsup{} variants compared with combinatory
superposition is that they rely on rules that enumerate possibly infinite sets
of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover
retains the same behavior as
before on first-order problems, smoothly scaling with
increasing numbers of higher-order clauses.
%
We also propose some heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).

Using first-order backends to finish the proof is common practice in
higher-order reasoning. Since \lsup{} coincides with standard
superposition on first-order clauses, invoking backends may
seem redundant; yet Zipperposition is nowhere as efficient as E
\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
efficient backend does make sense. We describe how to achieve a balance
between allowing native higher-order reasoning and
delegating reasoning to a backend (Sect.~\ref{sec:ho-tech:backends}).
%
Finally, we compare Zipperposition~2 with other provers on all monomorphic
higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
evaluation than at CASC (Sect.~\ref{sec:ho-tech:comparison}). Our evaluation
corroborates the competition results.

@@

\section{Background and Setting}
\label{sec:ho-tech:background}

@@

\section{Preprocessing Higher-Order Problems}
\label{sec:ho-tech:preprocessing}

@@

\section{Reasoning about Formulas}
\label{sec:ho-tech:formulas}

@@

\section{Exploring Boolean Selection Functions}
\label{sec:ho-tech:bool-select}

@@

When the input problem is clausified using immediate clausification, almost
all Boolean structure is lost. In this case, we expect Boolean selection to
have a modest effect. To better assess this feature, in this evaluation we use
\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:ho-tech:formulas} as the baseline
configuration. To avoid interference of literal and Boolean selection, we
additionally forbid the literal selection function from selecting a literal if
it contains a selectable Boolean subterm.

@@

\looseness=-1
On TPTP benchmarks, Boolean selection helps tame the explosion caused by
dynamic clausification used as inference: \NumberOK{All but one} selection functions
outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
\texttt{NUM636\^{}2} from Sect.~\ref{sec:ho-tech:formulas}, using Boolean selection
can reduce the number of given clause loop iterations from
\NumberOK{294} to \NumberOK{71}.

@@

\section{Enumerating Infinitely Branching Inferences}
\label{sec:ho-tech:infinite-branching}

@@

\section{Controlling Prolific Rules}
\label{sec:ho-tech:explosiveness}

@@

\section{Controlling the Use of Backends}
\label{sec:ho-tech:backends}

@@

\section{Comparison with Other Provers}
\label{sec:ho-tech:comparison}

%The raw evaluation data for the previous experiments show that
Different choices of
parameters lead to noticeably different sets of proved problems. In an attempt
to use Zipperposition~2 to its full potential, we have created a portfolio mode
that runs up to 50 configurations in parallel during the allotted time. The portfolio
was designed to solve as many problems as possible from the TPTP benchmark set. To
provide some context, we compare Zipperposition~2 with the latest versions of
all higher-order provers that competed at CASC-J10:\ CVC4 1.9
\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
The provers were run using the same parameters as in CASC, but with updated
executables.
Note that
Vampire's higher-order schedule is optimized for running on a single core.
We also include E~2.7 (more precisely, Ehoh, its higher-order configuration), the first version of this prover to syntactically support
full higher-order logic, including $\lambda$-abstractions.
Semantically, E~2.7 is arguably the weakest among the listed provers:
It simply performs $o$-RW rewriting described in Sect.~\ref{sec:ho-tech:preprocessing} followed by
$\lambda$-lifting before it applies \relax{$\lambda$-free superposition} \cite{bbcw-21-lfho}
(a precursor of all three higher-order superposition calculi) on the preprocessed problem.

@@

% \newcommand\ourparagraph[1]{\noindent{\textbf{\upshape#1.}}}
\newcommand\ourparagraph[1]{\subsection{#1}}

@@

\section{Introduction}
\label{sec:introduction}

@@

For many years, SAT solvers have used inexpensive clause simplification
techniques such as hidden literal and hidden tautology elimination
\cite{hjb-2010-cl-elim,hjb-2011-big-simplification} and failed literal
detection \cite[Sect.~1.6]{jwf-1995-fld}. We generalize these techniques to
first-order logic with equality
(Sect.~\ref{sec:hidden-literal-based-elimination}). Since the generalization
involves reasoning about infinite sets of literals, we propose restrictions to
make them usable.

@@

Specializing second-order quantifier elimination
\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
problems, yielding a technique we call \emph{singular predicate elimination}. We
extend their work along two axes (Sect.~\ref{sec:predicate-elimination}): We
generalize E\'en and Biere's refinement to first-order logic, resulting in
\emph{defined predicate elimination}, and explain how both types of predicate
elimination can be used during the proof search as inprocessing.

\begin{qle}
It is well known that clauses containing \emph{pure literals}---predicate
literals whose root symbol occurs only with one polarity in the problem---can
be eliminated. We propose a generalization of this technique, under the name
\emph{quasipure literal elimination}, to allow the elimination of some bipolar
literals as well (Sect.~\ref{sec:quasipure-literal-elimination}).
\end{qle}

The last technique we study is \emph{blocked clause elimination} (Sect.\
\ref{sec:satisfiability-by-clause-elimination}). It is used in both SAT
\cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
logic has produced good results when used as a preprocessor, especially on
satisfiable problems \cite{ksstb-2017-blockedfol}. We explore more ways to use
blocked clause elimination on satisfiable problems, including using it to
establish equisatisfiability with an empty clause set or as an inprocessing
rule. Unfortunately, we find that its use as inprocessing can compromise the
refutational completeness of superposition.

@@

% To improve the success rate of superposition provers, we extend several of
% preprocessing and inprocessing techniques from propositional to first-order logic,
% including failed literal detection, hidden literal elimination, hidden
% tautology elimination
% (Sect.~\ref{sec:hidden-literal-based-elimination}). These
% techniques are fairly inexpensive in propositional logic but need to be
% curtailed in first-order logic because they would otherwise involve reasoning
% about an infinite set of literals.

% Khasidashvili and Korovin \cite{kk-2016-pe-fol} generalized Davis--Putnam
% resolution to first-order logic with equality. We call their technique
% \emph{singular predicate elimination}. We also adapt bounded variable
% elimination by substitution, yielding \emph{defined predicate elimination}
% (Sect.~\ref{sec:predicate-elimination}). With some care, we can integrate these
% techniques with superposition without losing refutational completeness.

% Finally, we study a first-order generalization of blocked clause
% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satisfiability-by-clause-elimination}).
% %Unlike the other techniques, it weakens the clause set.
% If all clauses in the current clause set are blocked, this means that it is
% satisfiable.
% %It can be used to show
% %satisfiability by establishing equisatisfiability between the initial clause
% %set and the empty clause set.
% We show how this technique can help superposition provers detect
% satisfiability early instead of diverging. We also show, however, that
% interleaving the technique with superposition compromises refutational
% completeness.

All techniques are implemented in the Zipperposition prover
(Sect.~\ref{sec:implementation}), allowing us to ascertain their usefulness
(Sect.~\ref{sec:evaluation}). The best configuration solves \NumberOK{160}
additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
\cite{gs-17-tptp}. The raw experimental data are publicly available.%
\footnote{\url{https://doi.org/10.5281/zenodo.4552499}}\confrep{
More details, including all the proofs, can be found in a
technical report \cite{our-report}.}{}

@@

\section{Preliminaries}
\label{sec:preliminaries}

\ourparagraph{Clausal First-Order Logic}
\looseness=-1
Our setting is many-sorted, or many-typed, first-order logic \cite{jg-1987-logic-textbook} with
interpreted equality and a distinguished type (or sort) $o$.
Each variable $x$ is assigned a non-Boolean
type, and each symbol $\cst{f}$ is assigned a tuple $(\tau_1, \ldots, \tau_n,
\tau)$ where $n \geq 0$, $\tau_i$ are non-Boolean types, and $\tau$ is
the \emph{result type}. We distinguish between
\emph{predicate symbols}, with $o$ as the result type, and \emph{function symbols}.
Nullary function symbols are called \emph{constants}. Terms are either
variables~$x$ or \confrep{well-typed }{}applications $\cst{f}(t_1, \ldots,
t_n)$\confrep{, or $\cst{f}$ if $n = 0.$}{, of type $\tau$ where $\cst{f}$ is assigned $(\tau_1, \ldots,
\tau_n, \tau)$, and each $t_i$ is a term of type $\tau_i$.
The parentheses are omitted if $n = 0$.} A term is
\emph{ground} if it contains no variables.
We assume standard definitions and
notations for positions, subterms, and contexts \cite{bn-98-tr-and-all-that}.
We abbreviate a vector $(a_1, \ldots, a_n)$ to $\tuplen{a}$ or $\tuple{a}{}$,
and write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol $\cst{f}$
(e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$).

@@

\ourparagraph{Superposition Provers}
%
Superposition \cite{bg-94-superposition} is a calculus for clausal
first-order logic that extends ordered resolution
\cite{bg-01-resolution} with equality reasoning. It is
refutationally complete: Given a finite, unsatisfiable clause set, it will
eventually derive the empty clause. It is parameterized by a \emph{selection
function} that influences which of a clause's literals are eligible as the target
of inferences. Moreover, it is compatible with the \emph{standard redundancy
criterion}, which can be used to delete a clause $C$ while preserving completeness of the calculus.

@@

\section{Hidden-Literal-Based Elimination}
\label{sec:hidden-literal-based-elimination}

@@

\section{Predicate Elimination}
\label{sec:predicate-elimination}

@@

\begin{qle}
\section{Quasipure Literal Elimination}
\label{sec:quasipure-literal-elimination}

@@

Definition~\ref{def:qle} suggests a naive, nondeterminisitc procedure for
discovering and eliminating quasipure predicate symbols: Choose a predicate
symbol $\cst{p}$ and a polarity $s_\cst{p}$, and take $P = \{\cst{p}\}$. If the
predicate occurs with the wrong polarity in a clause
$\arbpredlit{\cst{p}(\ldots)} \lor \arbpredlit{\cst{q}_1(\ldots)} \lor \cdots \lor
\arbpredlit{\cst{q}_n(\ldots)} \lor C$, try to extend the polarity map $s$ and
the predicate set $P$ with one of the $\arbpredlit{\cst{q}_i}$'s and continue
recursively with $\cst{q}_i$. In Sect.~\ref{sec:implementation}, we will see a
more efficient approach based on a SAT encoding.
\end{qle}

\section{Satisfiability by Clause Elimination}
\label{sec:satisfiability-by-clause-elimination}

@@

\section{Implementation}
\label{sec:implementation}
% {\large\bf {TODO: Shorten and introduce SPPE here in a sentence}}

@@

\ourparagraph{Hidden-Literal-Based Elimination}
% 
% Heule et al.\ \cite{hjb-2010-cl-elim} show how hidden tautologies can be removed
% without creating the binary implication graph. For each literal~$L$ occurring in
% the clause set $N$, they compute the set $\PHL(L,N)$. Then for each $\neglit{L}'
% \in \PHL(L,N)$, they label $L'$ with $L$. A clause $C$ is a hidden tautology if
% there are two literals $L,L' \in C$ such that $L'$ is labeled with $L$. The
% described approach can be generalized to first-order logic and used to implement
% all rules from Sect.~\ref{sec:hidden-literal-based-elimination}.
% 
For HLBE, an efficient representation of $\HL(L,N)$ is crucial. Because
this set may be infinite, we underapproximate it by
restricting the length of the
transitive chains via a parameter $K_\mathrm{len}$. Given the current
clause set $N$, the finite map $\impmap[L']$ associates with each literal $L'$
a set of pairs $(L,M)$ such that $L' \bigfimpl{N}^k L$, where $k \leq
K_\mathrm{len}$ and $M$ is the multiset of clauses used to derive $L' \bigfimpl{N}^k L$.
Moreover, we consider only transitions of type (1) (as per
Definition~\ref{def:hl-fo}).
%Intuitively,
%$\impmap[L']$ contains all literals which can be reached from $L'$ following $k$
%edges of the generalized binary implication graph.
The following algorithm
maintains $\impmap$ dynamically, updating it as the prover derives and deletes
clauses.
It depends on the global variable $\impmap$ and the parameters
$K_\mathrm{len}$ and $K_{\mathrm{imp}}$.

@@

\ourparagraph{Predicate Elimination}
% 
To implement portfolio predicate elimination, we maintain a record for each
predicate symbol $\cst{p}$ occurring in the problem with the
following fields:\
set of definition clauses for $\cst{p}$, set of nondefinition clauses in which
$\cst{p}$ occurs once, and set of clauses in which $\cst{p}$ occurs more than
once. These records are kept in a priority queue, prioritized by properties such
as presence of definition sets and number of estimated resolutions. If
$\cst{p}$ is the highest-priority symbol that is eligible for SPE or DPE, we
eliminate it by removing all the clauses stored in $\cst{p}$'s record from the
proof state and by adding flat resolvents to the passive set.
Eliminating a symbol might make another symbol eligible.
%In this case, the updated record is put on the processing queue.

@@

\begin{qle}
\ourparagraph{Quasipure Literal Elimination}

@@

\section{Evaluation}
\label{sec:evaluation}

@@

\ourparagraph{Hidden-Literal-Based Elimination}
% 
The first experiments use all implemented HLBE rules.
To avoid overburdening Zipperposition, we can enable an option to limit the
number of tracked clauses for hidden literals. Once the limit has been reached, any
request for tracking a clause will be rejected until a tracked
clause is deleted. We can choose which kind of clauses are
tracked:\ only clauses from the active set $\mathcalx{A}$,
only clauses from the passive set $\mathcalx{P}$, or both.
We also vary the maximal implication chain length $K_\mathrm{len}$ and the
number of computed self-implications $K_\mathrm{imp}$.

@@

\ourparagraph{Predicate and Blocked Clause Elimination}
%
% \textit{Evaluation results discussion paragraph---here effects of definitions can be discussed}
% \textit{Do not forget to mention SPE was already implemented and }
For defined predicate elimination, the number of resolvents grows
exponentially with the number of occurrences of $\cst{p}$. To avoid this expensive
computation, we limit the applicability of PPE to proof states for which $\cst{p}$
is singular. According to our informal experiments, full PPE,
without this restriction, generally performs less well.

@@

\begin{qle}
\ourparagraph{Pure and Quasipure Literal Elimination}
\color{red}
Pure literal elimination is a known technique, but quasipure literals are new,
so it is quite interesting to find out whether they bring benefits on
benchmarks. Figure~\ref{ig:ple-qle} considers PLE, QLE, and a combination
of QLE and portfolio predicate elimination, applied in that order. All three
modes are tried both as preprocessing and as inprocessing.

@@

\ourparagraph{Satisfiability by Blocked Clause Elimination}
%
\looseness=-1%
Kiesl et al.\ found that blocked clause elimination is especially effective on
satisfiable problems. To corroborate their results and ascertain
whether a combination of predicate elimination and blocked clause elimination
increases the success rate, we evaluate BCE on all 2273
satisfiable or TPTP FOF and CNF problems.
The hardware and CPU time limits are the same as in the experiments above.
Figure~\ref{fig:bce-sat} presents the results.

@@

\begin{rep}
\section{Discussion and Related Work}
\label{sec:discussion-and-related-work}

@@

We briefly surveyed related work in Sect.~\ref{sec:introduction}. In this
section, we give a more detail overview and further discuss connections with
the related work.

@@

\section{Conclusion}
\label{sec:conclusion}

>>>>

\section{Introduction}
\label{sec:satfol:ho-tech:intro}

@@

Many higher-order problems extensively use symbol definitions to simplify
their representation. We describe several ways to exploit the definitions,
%of which the most successful is
such as turning them into rewrite rules (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
%Interesting patterns can be observed in various higher-order problem encodings.
%We show how we can exploit these to simplify problems (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
%
By working on formulas rather than clauses, tableau techniques take a more
holistic view of a higher-order problem.
Through its support for delayed clausification and, more generally,
calculus-level formula manipulation, \osup{} enables us to
simulate most successful tableau techniques in a saturating prover
(Sect.~\ref{sec:satfol:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
functions}, a mechanism that allows us to choose on which Boolean subterms
to perform inferences first.
We implemented some Boolean selection functions and
evaluated them (Sect.~\ref{sec:satfol:ho-tech:bool-select}).

The main drawback of both \lsup{} variants compared with combinatory
superposition is that they rely on rules that enumerate possibly infinite sets
of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:satfol:ho-tech:infinite-branching}). The prover
retains the same behavior as
before on first-order problems, smoothly scaling with
increasing numbers of higher-order clauses.
%
We also propose some heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:satfol:ho-tech:explosiveness}).

Using first-order backends to finish the proof is common practice in
higher-order reasoning. Since \lsup{} coincides with standard
superposition on first-order clauses, invoking backends may
seem redundant; yet Zipperposition is nowhere as efficient as E
\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
efficient backend does make sense. We describe how to achieve a balance
between allowing native higher-order reasoning and
delegating reasoning to a backend (Sect.~\ref{sec:satfol:ho-tech:backends}).
%
Finally, we compare Zipperposition~2 with other provers on all monomorphic
higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
evaluation than at CASC (Sect.~\ref{sec:satfol:ho-tech:comparison}). Our evaluation
corroborates the competition results.

@@

\section{Background and Setting}
\label{sec:satfol:ho-tech:background}

@@

\section{Preprocessing Higher-Order Problems}
\label{sec:satfol:ho-tech:preprocessing}

@@

\section{Reasoning about Formulas}
\label{sec:satfol:ho-tech:formulas}

@@

\section{Exploring Boolean Selection Functions}
\label{sec:satfol:ho-tech:bool-select}

@@

When the input problem is clausified using immediate clausification, almost
all Boolean structure is lost. In this case, we expect Boolean selection to
have a modest effect. To better assess this feature, in this evaluation we use
\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:satfol:ho-tech:formulas} as the baseline
configuration. To avoid interference of literal and Boolean selection, we
additionally forbid the literal selection function from selecting a literal if
it contains a selectable Boolean subterm.

@@

\looseness=-1
On TPTP benchmarks, Boolean selection helps tame the explosion caused by
dynamic clausification used as inference: \NumberOK{All but one} selection functions
outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
\texttt{NUM636\^{}2} from Sect.~\ref{sec:satfol:ho-tech:formulas}, using Boolean selection
can reduce the number of given clause loop iterations from
\NumberOK{294} to \NumberOK{71}.

@@

\section{Enumerating Infinitely Branching Inferences}
\label{sec:satfol:ho-tech:infinite-branching}

@@

\section{Controlling Prolific Rules}
\label{sec:satfol:ho-tech:explosiveness}

@@

\section{Controlling the Use of Backends}
\label{sec:satfol:ho-tech:backends}

@@

\section{Comparison with Other Provers}
\label{sec:satfol:ho-tech:comparison}

%The raw evaluation data for the previous experiments show that
Different choices of
parameters lead to noticeably different sets of proved problems. In an attempt
to use Zipperposition~2 to its full potential, we have created a portfolio mode
that runs up to 50 configurations in parallel during the allotted time. The portfolio
was designed to solve as many problems as possible from the TPTP benchmark set. To
provide some context, we compare Zipperposition~2 with the latest versions of
all higher-order provers that competed at CASC-J10:\ CVC4 1.9
\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
The provers were run using the same parameters as in CASC, but with updated
executables.
Note that
Vampire's higher-order schedule is optimized for running on a single core.
We also include E~2.7 (more precisely, Ehoh, its higher-order configuration), the first version of this prover to syntactically support
full higher-order logic, including $\lambda$-abstractions.
Semantically, E~2.7 is arguably the weakest among the listed provers:
It simply performs $o$-RW rewriting described in Sect.~\ref{sec:satfol:ho-tech:preprocessing} followed by
$\lambda$-lifting before it applies \relax{$\lambda$-free superposition} \cite{bbcw-21-lfho}
(a precursor of all three higher-order superposition calculi) on the preprocessed problem.

@@

% \newcommand\ourpara[1]{\noindent{\textbf{\upshape#1.}}}
\newcommand\ourpara[1]{\subsection{#1}}

@@

\section{Introduction}
\label{sec:satfol:introduction}

@@

For many years, SAT solvers have used inexpensive clause simplification
techniques such as hidden literal and hidden tautology elimination
\cite{hjb-2010-cl-elim,hjb-2011-big-simplification} and failed literal
detection \cite[Sect.~1.6]{jwf-1995-fld}. We generalize these techniques to
first-order logic with equality
(Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). Since the generalization
involves reasoning about infinite sets of literals, we propose restrictions to
make them usable.

@@

Specializing second-order quantifier elimination
\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
problems, yielding a technique we call \emph{singular predicate elimination}. We
extend their work along two axes (Sect.~\ref{sec:satfol:predicate-elimination}): We
generalize E\'en and Biere's refinement to first-order logic, resulting in
\emph{defined predicate elimination}, and explain how both types of predicate
elimination can be used during the proof search as inprocessing.

\begin{qle}
It is well known that clauses containing \emph{pure literals}---predicate
literals whose root symbol occurs only with one polarity in the problem---can
be eliminated. We propose a generalization of this technique, under the name
\emph{quasipure literal elimination}, to allow the elimination of some bipolar
literals as well (Sect.~\ref{sec:satfol:quasipure-literal-elimination}).
\end{qle}

The last technique we study is \emph{blocked clause elimination} (Sect.\
\ref{sec:satfol:satisfiability-by-clause-elimination}). It is used in both SAT
\cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
logic has produced good results when used as a preprocessor, especially on
satisfiable problems \cite{ksstb-2017-blockedfol}. We explore more ways to use
blocked clause elimination on satisfiable problems, including using it to
establish equisatisfiability with an empty clause set or as an inprocessing
rule. Unfortunately, we find that its use as inprocessing can compromise the
refutational completeness of superposition.

@@

% To improve the success rate of superposition provers, we extend several of
% preprocessing and inprocessing techniques from propositional to first-order logic,
% including failed literal detection, hidden literal elimination, hidden
% tautology elimination
% (Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). These
% techniques are fairly inexpensive in propositional logic but need to be
% curtailed in first-order logic because they would otherwise involve reasoning
% about an infinite set of literals.

% Khasidashvili and Korovin \cite{kk-2016-pe-fol} generalized Davis--Putnam
% resolution to first-order logic with equality. We call their technique
% \emph{singular predicate elimination}. We also adapt bounded variable
% elimination by substitution, yielding \emph{defined predicate elimination}
% (Sect.~\ref{sec:satfol:predicate-elimination}). With some care, we can integrate these
% techniques with superposition without losing refutational completeness.

% Finally, we study a first-order generalization of blocked clause
% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satfol:satisfiability-by-clause-elimination}).
% %Unlike the other techniques, it weakens the clause set.
% If all clauses in the current clause set are blocked, this means that it is
% satisfiable.
% %It can be used to show
% %satisfiability by establishing equisatisfiability between the initial clause
% %set and the empty clause set.
% We show how this technique can help superposition provers detect
% satisfiability early instead of diverging. We also show, however, that
% interleaving the technique with superposition compromises refutational
% completeness.

All techniques are implemented in the Zipperposition prover
(Sect.~\ref{sec:satfol:implementation}), allowing us to ascertain their usefulness
(Sect.~\ref{sec:satfol:evaluation}). The best configuration solves \NumberOK{160}
additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
\cite{gs-17-tptp}. The raw experimental data are publicly available.%
\footnote{\url{https://doi.org/10.5281/zenodo.4552499}}\confrep{
More details, including all the proofs, can be found in a
technical report \cite{our-report}.}{}

@@

\section{Preliminaries}
\label{sec:satfol:preliminaries}

\ourpara{Clausal First-Order Logic}
\looseness=-1
Our setting is many-sorted, or many-typed, first-order logic \cite{jg-1987-logic-textbook} with
interpreted equality and a distinguished type (or sort) $o$.
Each variable $x$ is assigned a non-Boolean
type, and each symbol $\cst{f}$ is assigned a tuple $(\tau_1, \ldots, \tau_n,
\tau)$ where $n \geq 0$, $\tau_i$ are non-Boolean types, and $\tau$ is
the \emph{result type}. We distinguish between
\emph{predicate symbols}, with $o$ as the result type, and \emph{function symbols}.
Nullary function symbols are called \emph{constants}. Terms are either
variables~$x$ or \confrep{well-typed }{}applications $\cst{f}(t_1, \ldots,
t_n)$\confrep{, or $\cst{f}$ if $n = 0.$}{, of type $\tau$ where $\cst{f}$ is assigned $(\tau_1, \ldots,
\tau_n, \tau)$, and each $t_i$ is a term of type $\tau_i$.
The parentheses are omitted if $n = 0$.} A term is
\emph{ground} if it contains no variables.
We assume standard definitions and
notations for positions, subterms, and contexts \cite{bn-98-tr-and-all-that}.
We abbreviate a vector $(a_1, \ldots, a_n)$ to $\tuplen{a}$ or $\tuple{a}{}$,
and write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol $\cst{f}$
(e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$).

@@

\ourpara{Superposition Provers}
%
Superposition \cite{bg-94-superposition} is a calculus for clausal
first-order logic that extends ordered resolution
\cite{bg-01-resolution} with equality reasoning. It is
refutationally complete: Given a finite, unsatisfiable clause set, it will
eventually derive the empty clause. It is parameterized by a \emph{selection
function} that influences which of a clause's literals are eligible as the target
of inferences. Moreover, it is compatible with the \emph{standard redundancy
criterion}, which can be used to delete a clause $C$ while preserving completeness of the calculus.

@@

\section{Hidden-Literal-Based Elimination}
\label{sec:satfol:hidden-literal-based-elimination}

@@

\section{Predicate Elimination}
\label{sec:satfol:predicate-elimination}

@@

\begin{qle}
\section{Quasipure Literal Elimination}
\label{sec:satfol:quasipure-literal-elimination}

@@

Definition~\ref{def:qle} suggests a naive, nondeterminisitc procedure for
discovering and eliminating quasipure predicate symbols: Choose a predicate
symbol $\cst{p}$ and a polarity $s_\cst{p}$, and take $P = \{\cst{p}\}$. If the
predicate occurs with the wrong polarity in a clause
$\arbpredlit{\cst{p}(\ldots)} \lor \arbpredlit{\cst{q}_1(\ldots)} \lor \cdots \lor
\arbpredlit{\cst{q}_n(\ldots)} \lor C$, try to extend the polarity map $s$ and
the predicate set $P$ with one of the $\arbpredlit{\cst{q}_i}$'s and continue
recursively with $\cst{q}_i$. In Sect.~\ref{sec:satfol:implementation}, we will see a
more efficient approach based on a SAT encoding.
\end{qle}

\section{Satisfiability by Clause Elimination}
\label{sec:satfol:satisfiability-by-clause-elimination}

@@

\section{Implementation}
\label{sec:satfol:implementation}
% {\large\bf {TODO: Shorten and introduce SPPE here in a sentence}}

@@

\ourpara{Hidden-Literal-Based Elimination}
% 
% Heule et al.\ \cite{hjb-2010-cl-elim} show how hidden tautologies can be removed
% without creating the binary implication graph. For each literal~$L$ occurring in
% the clause set $N$, they compute the set $\PHL(L,N)$. Then for each $\neglit{L}'
% \in \PHL(L,N)$, they label $L'$ with $L$. A clause $C$ is a hidden tautology if
% there are two literals $L,L' \in C$ such that $L'$ is labeled with $L$. The
% described approach can be generalized to first-order logic and used to implement
% all rules from Sect.~\ref{sec:satfol:hidden-literal-based-elimination}.
% 
For HLBE, an efficient representation of $\HL(L,N)$ is crucial. Because
this set may be infinite, we underapproximate it by
restricting the length of the
transitive chains via a parameter $K_\mathrm{len}$. Given the current
clause set $N$, the finite map $\impmap[L']$ associates with each literal $L'$
a set of pairs $(L,M)$ such that $L' \bigfimpl{N}^k L$, where $k \leq
K_\mathrm{len}$ and $M$ is the multiset of clauses used to derive $L' \bigfimpl{N}^k L$.
Moreover, we consider only transitions of type (1) (as per
Definition~\ref{def:hl-fo}).
%Intuitively,
%$\impmap[L']$ contains all literals which can be reached from $L'$ following $k$
%edges of the generalized binary implication graph.
The following algorithm
maintains $\impmap$ dynamically, updating it as the prover derives and deletes
clauses.
It depends on the global variable $\impmap$ and the parameters
$K_\mathrm{len}$ and $K_{\mathrm{imp}}$.

@@

\ourpara{Predicate Elimination}
% 
To implement portfolio predicate elimination, we maintain a record for each
predicate symbol $\cst{p}$ occurring in the problem with the
following fields:\
set of definition clauses for $\cst{p}$, set of nondefinition clauses in which
$\cst{p}$ occurs once, and set of clauses in which $\cst{p}$ occurs more than
once. These records are kept in a priority queue, prioritized by properties such
as presence of definition sets and number of estimated resolutions. If
$\cst{p}$ is the highest-priority symbol that is eligible for SPE or DPE, we
eliminate it by removing all the clauses stored in $\cst{p}$'s record from the
proof state and by adding flat resolvents to the passive set.
Eliminating a symbol might make another symbol eligible.
%In this case, the updated record is put on the processing queue.

@@

\begin{qle}
\ourpara{Quasipure Literal Elimination}

@@

\section{Evaluation}
\label{sec:satfol:evaluation}

@@

\ourpara{Hidden-Literal-Based Elimination}
% 
The first experiments use all implemented HLBE rules.
To avoid overburdening Zipperposition, we can enable an option to limit the
number of tracked clauses for hidden literals. Once the limit has been reached, any
request for tracking a clause will be rejected until a tracked
clause is deleted. We can choose which kind of clauses are
tracked:\ only clauses from the active set $\mathcalx{A}$,
only clauses from the passive set $\mathcalx{P}$, or both.
We also vary the maximal implication chain length $K_\mathrm{len}$ and the
number of computed self-implications $K_\mathrm{imp}$.

@@

\ourpara{Predicate and Blocked Clause Elimination}
%
% \textit{Evaluation results discussion paragraph---here effects of definitions can be discussed}
% \textit{Do not forget to mention SPE was already implemented and }
For defined predicate elimination, the number of resolvents grows
exponentially with the number of occurrences of $\cst{p}$. To avoid this expensive
computation, we limit the applicability of PPE to proof states for which $\cst{p}$
is singular. According to our informal experiments, full PPE,
without this restriction, generally performs less well.

@@

\begin{qle}
\ourpara{Pure and Quasipure Literal Elimination}
\color{red}
Pure literal elimination is a known technique, but quasipure literals are new,
so it is quite interesting to find out whether they bring benefits on
benchmarks. Figure~\ref{ig:ple-qle} considers PLE, QLE, and a combination
of QLE and portfolio predicate elimination, applied in that order. All three
modes are tried both as preprocessing and as inprocessing.

@@

\ourpara{Satisfiability by Blocked Clause Elimination}
%
\looseness=-1%
Kiesl et al.\ found that blocked clause elimination is especially effective on
satisfiable problems. To corroborate their results and ascertain
whether a combination of predicate elimination and blocked clause elimination
increases the success rate, we evaluate BCE on all 2273
satisfiable or TPTP FOF and CNF problems.
The hardware and CPU time limits are the same as in the experiments above.
Figure~\ref{fig:bce-sat} presents the results.

@@

\begin{rep}
\section{Discussion and Related Work}
\label{sec:satfol:discussion-and-related-work}

@@

We briefly surveyed related work in Sect.~\ref{sec:satfol:introduction}. In this
section, we give a more detail overview and further discuss connections with
the related work.

@@

\section{Conclusion}
\label{sec:satfol:conclusion}
